---
title: "Project Proposal"
author: "Kossi Akplaka"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
```

Let's load the data in a dataframe

```{r load-data}
Complaint_data <- read.csv("https://raw.githubusercontent.com/Kossi-Akplaka/Data606-Stats_and_Probability_in_R/main/Project%20Proposal/NYPD_Complaint_Data_Historic.csv")

kable(Complaint_data)
```
Data exploration 

Let's have a glimpse of the data. 
```{r}
glimpse(Complaint_data)
```


We can look up the missing values. As a reminder, the missing values in this data is recorded as (null)

```{r}
# Look up the missing values
missing_data <- Complaint_data %>%
  summarise_all(~ sum(.=="(null)"))

kable(missing_data)
```
This data frame has a lot of columns. We can drop the column that has a lot of missing values and the columns we are going to use for the analysis

```{r}
Complaint_data <- Complaint_data %>% 
  select(- CMPLNT_NUM,  # Unique ID of the records
         - PARKS_NM,
         - HADEVELOPT, 
         - HOUSING_PSA,
         - STATION_NAME, 
         -LOC_OF_OCCUR_DESC,
         - Latitude, - Longitude, - Lat_Lon)  # A lot of missing values

kable(Complaint_data)
```




### Research question 

- Do law enforcement agencies tend to respond more quickly to reported incidents of "DANGEROUS WEAPONS" compared to "DANGEROUS DRUGS"? 

- Do law enforcement agencies tend to respond more quickly to reported incidents of "DANGEROUS WEAPONS" in Queens Vs Manhattan 

- 



### Cases 

**What are the cases, and how many are there?**



### Data collection 

**Describe the method of data collection.**



### Type of study 

**What type of study is this (observational/experiment)?**



### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**




### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**



### Independent Variable(s)




### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

```
